The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting

### 2024-12-13 manuscript/01-introduction.qmd
final draft for submission

Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors. These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding. [@schneider2010pharmaceutical; @hyndman2018forecasting]. Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights. This investment is essential and offers significant benefits for improving supply chain delivery. However, much of the data being collected focuses on transactional data, such as consumption and sale. Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting. This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models. This gap—known as business context and domain knowledge—requires urgent attention in Pharmaceutical supply chain forecasting. Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results. We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains. Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview]. Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model. This integration is particularly beneficial in fields like healthcare, where expert insights can align algorithms with real-world applications [@dash2022review].

@@

This research uses consumption data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting. Initially, we implement univariate models that rely solely on consumption data. Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns. Through these interviews, the experts identified five key factors collectively impacting consumption: stock replenishment period, stockouts, fiscal calendar, COVID-19, conflicts, and the seasonal malaria. We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.

Our primary aims is to assess whether incorporating domain knowledge improves forecast accuracy. Moreover, we also fill a gap in the pharmaceutical demand forecasting literature by not only generating point forecasts but also producing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods. Model performance will be evaluated using point accuracy measure including Mean Absolute Scaled Error(MASE) and Root Mean Squared Scaled Error (RMSSE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.

>>>>

Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors. These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding [@schneider2010pharmaceutical; @hyndman2018forecasting]. Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights. This investment is essential and offers significant benefits for improving supply chain delivery. However, much of the data being collected focuses on transactional data, such as consumption and sale. Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting. This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models. This gap—known as business context and domain knowledge—requires urgent attention in Pharmaceutical supply chain forecasting. Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results. We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains. Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview]. Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model. This integration is particularly beneficial in fields like healthcare, where expert insights can align algorithms with real-world applications [@dash2022review].

@@

This research uses consumption data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting. Initially, we implement univariate models that rely solely on consumption data. Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns. Through these interviews, the experts identified three key factors collectively impacting consumption: stock replenishment period, fiscal calendar, and the seasonal malaria. We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.

Our primary aims is to assess whether incorporating domain knowledge improves forecast accuracy. Moreover, we also fill a gap in the pharmaceutical demand forecasting literature by not only generating point forecasts but also producing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods. Model performance is evaluated using point accuracy measure including MASE, RMSSE, and probabilistic measure using CRPS.

### 2024-12-13 manuscript/02-literature.qmd
final draft for submission

Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes. Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed. The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid, @makridakis2020forecasting].

In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting,@zhu2021demand,@anusha2014demand]. Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications. In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand,@candan2014demand, and @ribeiro2017improving].

@@

The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques. These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.@van2021using, showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy. Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE. Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting. The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories. This suggests that no single model is universally optimal, and model selection should be context-specific.

>>>>

Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes. Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed. The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid; @makridakis2020forecasting].

In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand]. Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications. In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand, @candan2014demand, and @ribeiro2017improving.

@@

The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques. These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power. @van2021using showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy. Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE. Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting. The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories. This suggests that no single model is universally optimal, and model selection should be context-specific.

### 2024-12-13 manuscript/03-experiment.qmd
final draft for submission

To gather domain knowledge about factors affected the consumption of products, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit. Next, we collaborated with six experts from the Ethiopian Pharmaceuticals Supply Service (EPSS), each with over 10 years of experience managing pharmaceutical distribution. In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies. The experts provided insights based on their extensive experience and knowledge of distribution practices. When there was uncertainty or gaps in the data for certain products, additional experts were consulted to gather further context. If all experts reached a consensus, the influencing factors were confirmed and included in the analysis.

@@

- Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters. Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises. Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas. This disruption can cause significant delays or prevent the delivery of essential medicines altogether. Internal displacements and conflicts in Ethiopia over the past five years also affected consumption and distribution. We traced periods of instability and included them as predictors. The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.

@@

**ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors. ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model. This modification allows to include relevant information from external factors such as malaria season, and fiscal year period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_. By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models. In our study, we use an automated algorithm to determine the optimal configuration for ARIMA components, following the approach outlined by @hyndman2021forecasting and We use the ARIMA() function from the fable package in R.

@@

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@gneiting2014probabilistic;@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
$$
  \text{CRPS} = \text{mean}(p_j),
$$ {#eq-CRPS}
where
$$
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
$$
where $G_j(x)$ is the forecasted probability distribution function for forecast horizon $j$, and $F_j(x)$ is the true probability distribution function for the same period.

@@

CRPS can be considered an average of all possible qiantiles [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles. A specific prediction interval could be evaluated using a Winkler score, if required.

>>>>

To gather domain knowledge about factors affected the consumption of products, we contacted experts from EPSS who work in the distribution unit. Next, we collaborated with six experts, each with over 10 years of experience managing pharmaceutical distribution. In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies. The experts provided insights based on their extensive experience and knowledge of distribution practices. When there was uncertainty or gaps in the data for certain products, additional experts were consulted to gather further context. If all experts reached a consensus, the influencing factors were confirmed and included in the analysis.

@@

<!-- - Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters. Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises. Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas. This disruption can cause significant delays or prevent the delivery of essential medicines altogether. Internal displacements and conflicts in Ethiopia over the past five years also affected consumption and distribution. We traced periods of instability and included them as predictors. The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.  -->

@@

**ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors. ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model. This modification allows to include relevant information from external factors such as malaria season, and fiscal year period, and stock replenishement period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_. By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models. In our study, we use an automated algorithm to determine the optimal configuration for ARIMA components, following the approach outlined by @hyndman2021forecasting and We use the ARIMA() function from the fable package in R.

@@

In all models, probabilistic forecasts were generated using bootstrapping to create 1,000 possible future scenarios for each period within the forecast horizon [@hyndman2021forecasting].

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
$$
  \text{CRPS} = \text{mean}(p_j),
$$ {#eq-CRPS}
where
$$
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
$$
where $G_j(x)$ is the forecasted probability distribution function for forecast horizon $j$, and $F_j(x)$ is the true probability distribution function for the same period.

@@

CRPS can be considered an average of all possible quantiles [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles. A specific prediction interval could be evaluated using a Winkler score, if required.

### 2024-12-13 manuscript/04-results.qmd
final draft for submission

In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical consumption data. Point forecast performance is reported using MASE and MSSE, while probabilistic forecast accuracy is reported using CRPS.

The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin. we also report the ditribution of accurcy metrics  across all rollog origins, average across forecast horizon and all products. this shows how mdoels varies in roviding accuracy across different origins. further rserach would be required to understand how error analysis across forecast hrizon and across differne origins could be used as afiagnos!
y-axis shows the models sorted based on ther values of mase and rmsse, when the one i the bottom has lowest error. this modesis TimeGpt, which is the foundational time sei models.

@fig-point also indicate that predictors collected through interacting with experts with domaine knoeldge improves point forecast accuracy for all models, highlighting the essential needs to systematically colelct such information alognside the tarsactional data about consuption.

We observed that the LSTM model had the highest error compared to all other models. While LSTMs are powerful for capturing complex, non-linear temporal patterns, they may underperform in certain scenarios, especially when compared to simpler time series forecasting models. This underperformance can be attributed to overfitting, as LSTMs have a large number of parameters and require substantial amounts of data to generalize effectively. In contrast, simpler models like ARIMA with exogenous variables often handle straightforward trends or seasonality more effectively and may outperform LSTMs when long-term dependencies or non-linearity in the data are not significant.

@@

```{r}
#| label: fig-horizon
#| fig-width: 11
#| fig-height: 5
#| fig-cap: "Average accuracy by month for 6 month. The total number of months used to calculate the accuracy in the test set is 12 for each product."
#| fig-pos: "H"

p1h <- point_accuracy_h_all |> group_by(model,h) |> summarise(mase=mean(mase) , .groups = "drop")|> filter(model != "lstm", model != "lstm_reg") |>  ggplot(aes(x = h, y = mase, colour =model))+
  geom_point()+
  geom_line()+
  ggthemes::theme_few()+
  ggthemes::scale_color_colorblind()+
  labs(x= " Forecast horizon")+
  theme(legend.position="none")

p2h <- point_accuracy_h_all |> group_by(model,h) |> summarise(rmsse=mean(rmsse),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
  ggplot(aes(x = h, y = rmsse, colour =model))+
  geom_point()+
  geom_line()+
  ggthemes::theme_few()+
  ggthemes::scale_color_colorblind()+
  labs(x= " Forecast horizon")+
  theme(legend.position="bottom")

p3h <- prob_accuracy_h_all |> group_by(model,h) |> summarise(crps=mean(crps),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
  ggplot(aes(x = h, y = crps, colour =model))+
  geom_point()+
  geom_line()+
  ggthemes::theme_few()+
  ggthemes::scale_color_colorblind()+
  labs(x= " Forecast horizon")+
  theme(legend.position="none")

prob_patch <- (p1h|p2h|p3h)
prob_patch
```

@fig-prob illustrates the forecast distribution accuracy measured by the Continuous Ranked Probability Score (CRPS), which evaluates both forecasting calibration and interval sharpness. A smaller CRPS value indicates better overall performance. We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts. Notably, TimeGpt achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest. This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.

In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon. this is the accuracy reorted as the average across rolling origins and all products fro each given horizon. For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.

Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.

Our analysis highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors. In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.

The findings also emphasize the importance of incorporating relevant domain knowledge into forecasting models. In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on consumption and distribution. Including exogenous variables such as administrative procedures, seasonal patterns, or conflicts, or any other relevant factor that could be identified by those with domain knowledge proved essential for reducing forecast errors. This approach improved both point and probabilistic forecast accuracy, enabling a more confident assessment of uncertainty. Therefore, the systematic collection of information about significant events and their impact on consumption is vital. Recording details of events such as policy changes, administration procedures, conflicts, and incorporating this information into forecasting models creates a comprehensive understanding of consumption. This practice enhances modeling reliability and allows institutions in developing countries and humanitarian organizations to better forecast demand, allocate resources effectively, and respond proactively. Establishing robust data collection systems is thus a critical step for strengthening forecasting capabilities and operational resilience.

@@

fcst <- fit |> forecast(h = 6)
fitted_ets <- fit |> augment()

>>>>

In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical consumption data. Point forecast performance is reported using MASE and RMSSE, while probabilistic forecast accuracy is reported using CRPS.

The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin. We report the distribution of accuracy metrics across all rolling origins. This shows how models varies in providing accuracy across different origins. The y-axis displays models sorted by their MASE and RMSSE values, with the model exhibiting the lowest error positioned at the bottom. This model is the Ariam model, which incorporates exogenous variables. Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models. This underscores the critical importance of systematically collecting such expert-informed data alongside transactional consumption data.

@@

<!-- ```{r} -->
<!-- #| label: fig-horizon -->
<!-- #| fig-width: 11 -->
<!-- #| fig-height: 5 -->
<!-- #| fig-cap: "Average accuracy by month for 6 month. The total number of months used to calculate the accuracy in the test set is 12 for each product." -->
<!-- #| fig-pos: "H" -->

<!-- p1h <- point_accuracy_h_all |> group_by(model,h) |> summarise(mase=mean(mase) , .groups = "drop")|> filter(model != "lstm", model != "lstm_reg") |>  ggplot(aes(x = h, y = mase, colour =model))+ -->
<!--   geom_point()+ -->
<!--   geom_line()+ -->
<!--   ggthemes::theme_few()+ -->
<!--   ggthemes::scale_color_colorblind()+ -->
<!--   labs(x= " Forecast horizon")+ -->
<!--   theme(legend.position="none") -->

<!-- p2h <- point_accuracy_h_all |> group_by(model,h) |> summarise(rmsse=mean(rmsse),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |>  -->
<!--   ggplot(aes(x = h, y = rmsse, colour =model))+ -->
<!--   geom_point()+ -->
<!--   geom_line()+ -->
<!--   ggthemes::theme_few()+ -->
<!--   ggthemes::scale_color_colorblind()+ -->
<!--   labs(x= " Forecast horizon")+ -->
<!--   theme(legend.position="bottom") -->

<!-- p3h <- prob_accuracy_h_all |> group_by(model,h) |> summarise(crps=mean(crps),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |>  -->
<!--   ggplot(aes(x = h, y = crps, colour =model))+ -->
<!--   geom_point()+ -->
<!--   geom_line()+ -->
<!--   ggthemes::theme_few()+ -->
<!--   ggthemes::scale_color_colorblind()+ -->
<!--   labs(x= " Forecast horizon")+ -->
<!--   theme(legend.position="none")  -->

<!-- prob_patch <- (p1h|p2h|p3h) -->
<!-- prob_patch -->
<!-- ``` -->

@fig-prob illustrates the forecast distribution accuracy measured by CRPS, which evaluates both forecasting calibration and interval sharpness. A smaller CRPS value indicates better overall performance. We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts. Notably, ARIMA achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest. This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.

We observe that LSTM models exhibit a widespread error distribution compared to other models. This is attributed to their performance variability, which can significantly depend on the characteristics of demand patterns across different products. For example, the product "Amlodipine - 5mg - Tablet" demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero demand. Such patterns align well with the strengths of LSTM models, which excel at capturing long-term dependencies and managing complex temporal fluctuations.
In contrast, the demand for "Anti-Rho (D)" is erratic and sporadic, characterized by frequent, random fluctuations without any discernible long-term trends or seasonal patterns. This chaotic nature, combined with inconsistent and often low volumes, poses significant challenges for LSTM models. The lack of clear structure or stable patterns limits the models' ability to generalize effectively, resulting in poorer forecast performance. Although LSTM models can handle irregular data to some extent, they require sufficient and well-distributed training samples to achieve optimal performance. In our study, the majority of products exhibit erratic and sporadic demand patterns similar to "Anti-Rho (D)", resulting in this widespread error distribution.

<!-- In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon. this is the accuracy reorted as the average across rolling origins and all products fro each given horizon. For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.  -->

<!-- Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.  -->

The findings emphasize the importance of incorporating relevant domain knowledge into forecasting models. In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on consumption and distribution. Including exogenous variables such as administrative procedures, seasonal patterns, or conflicts, or any other relevant factor that could be identified by those with domain knowledge proved essential for reducing forecast errors. This approach improved both point and probabilistic forecast accuracy, enabling a more confident assessment of uncertainty. Therefore, the systematic collection of information about significant events and their impact on consumption is vital. Recording details of events such as policy changes, administration procedures, conflicts, and incorporating this information into forecasting models creates a comprehensive understanding of consumption. This practice enhances modeling reliability and allows institutions in developing countries and humanitarian organizations to better forecast demand, allocate resources effectively, and respond proactively. Establishing robust data collection systems is thus a critical step for strengthening forecasting capabilities and operational resilience.

Our analysis also highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors. In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.

@@

fcst <- fit |> forecast(h = 6) |> mutate(quantity = distributional::dist_truncated(quantity, lower = 0), .mean=mean(quantity))

fitted_ets <- fit |> augment()

### 2024-12-13 manuscript/05-conclusion.qmd
final draft for submission

In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the Ethiopian Pharmaceutical Supply Service (EPSS). Using five years of consumption data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to contribute to enhancing the demand forecasting process and emphasize the importance of integrating domain knowledge into model building. This step is critical, as the phase of understanding the data and incorporating valuable contextual information is often overlooked. Too frequently, forecasting efforts rely solely on consumption data, leading to building models with little relevance to reality. Our approach highlights the necessity of including domain insights to construct more accurate and effective forecasting models.

@@

>>>>

In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the EPSS. Using five years of consumption data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to contribute to enhancing the demand forecasting process and emphasize the importance of integrating domain knowledge into model building. This step is critical, as the phase of understanding the data and incorporating valuable contextual information is often overlooked. Too frequently, forecasting efforts rely solely on consumption data, leading to building models with little relevance to reality. Our approach highlights the necessity of including domain insights to construct more accurate and effective forecasting models.

@@

- Large Language Models (LLMs) for time series forecasting, like the one included in this study, offer transformative opportunities for developing countries by making advanced forecasting methods accessible, even in contexts with limited expertise or infrastructure. However, future research should explore their applicability, advantages, limitations, and performance relative to traditional and hybrid forecasting methods.

### 2024-12-13 manuscript/main.qmd
final draft for submission

---
title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
author:
  - name: Arebu Issa Bilal
    email: arebu.issa@aau.edu.et
    affiliations:
        - id: Addis 
          name: Addis Ababa University
          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
          country: Ethiopia.
    attributes:
        corresponding: true
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: Cardiff University
          department: Data Lab for Social Good Reserach Group, Cardiff Business School 
          country: UK.
  - name: Harsha Halgamuwe Hewage
    email: HalgamuweHewageHR@cardiff.ac.uk
    affiliations:
        - ref: carbs
  - name: Umit Sezer Bititci 
    email: U.S.Bititci@hw.ac.uk
    affiliations:
        - id: herio 
          name: Heriot-Watt University
          department: Edinburgh Business School, Heriot-Watt University
          country: UK.
  - name: Teferi Gedif Fenta
    email: tgedif@gmail.com
    affiliations:
        - ref: Addis

abstract: |
  Accurate forecasting in pharmaceutical supply chains is critical for ensuring the continuous availability of essential medicines, particularly in developing countries where resource constraints and logistical challenges are more pronounced. Effective demand forecasting supports procurement and inventory management, which in turn could help to prevent stockouts, reduce wastage due to overstocking, and enhance overall healthcare delivery. In such settings, reliable forecasts, that acknowledge uncertainties, are essential to ensure that limited resources are used efficiently to meet health needs. 
  Despite advancements in forecasting, significant gaps remain in effective forecasting in poor resource countries. Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption Furthermore, many works lack rigorous methodological design, transparency, and reproducibility. This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models. Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, conflicts, and disease outbreaks. Evaluation metrics such as Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy. The findings underscore the significance of contextual data in developing robust forecasting models that are adaptable to complex, real-world conditions. Our results also highlight the effectiveness of foundational time series methods, such as LLM models, for forecasting. These methods are particularly appealing for resource-constrained countries that may lack advanced analytical expertise. To promote usability, generalizability, and reproducibility, we share the complete dataset and code in R and Python, and the entire paper is written in Quarto via a GitHub repository to facilitate these practices.

keywords:
  - Forecasting
  - Pharmaceutical supply chain
  - Domain knowlege
  - Forecast accuracy
  - Developing countries
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  pdf-engine: pdflatex
  prefer-html: true
  cache: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: Journal of 
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

>>>>

---
title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
# author:
#   - name: Arebu Issa Bilal
#     email: arebu.issa@aau.edu.et
#     affiliations:
#         - id: Addis 
#           name: Addis Ababa University
#           department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
#           country: Ethiopia.
#     attributes:
#         corresponding: true
#   - name: Bahman Rostami-Tabar
#     email: rostami-tabarb@cardiff.ac.uk
#     affiliations:
#         - id: carbs
#           name: Cardiff University
#           department: Data Lab for Social Good Reserach Group, Cardiff Business School 
#           country: UK.
#   - name: Harsha Halgamuwe Hewage
#     email: HalgamuweHewageHR@cardiff.ac.uk
#     affiliations:
#         - ref: carbs
#   - name: Umit Sezer Bititci 
#     email: U.S.Bititci@hw.ac.uk
#     affiliations:
#         - id: herio 
#           name: Heriot-Watt University
#           department: Edinburgh Business School, Heriot-Watt University
#           country: UK.
#   - name: Teferi Gedif Fenta
#     email: tgedif@gmail.com
#     affiliations:
#         - ref: Addis

abstract: |
  Accurate forecasting in pharmaceutical supply chains is critical for ensuring the continuous availability of essential medicines, particularly in developing countries where resource constraints and logistical challenges are more pronounced. Effective demand forecasting supports procurement and inventory management, which in turn could help to prevent stockouts, reduce wastage due to overstocking, and enhance overall healthcare delivery. In such settings, reliable forecasts, that acknowledge uncertainties, are essential to ensure that limited resources are used efficiently to meet health needs. 
  Despite advancements in forecasting, significant gaps remain in effective forecasting in poor resource countries. Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts. Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption. Furthermore, many works lack rigorous methodological design, transparency, and reproducibility. This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models. Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, and disease outbreaks. Evaluation metrics including Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy. The findings underscore the significance of contextual data in developing robust forecasting models that are adaptable to complex, real-world conditions. Our results also highlight the effectiveness of foundational time series models for forecasting. These models are particularly appealing for resource-constrained countries that may lack advanced analytical expertise. To promote usability, generalizability, and reproducibility, we share the complete dataset and code in R and Python, and the entire paper is written in Quarto via a GitHub repository to facilitate these practices.

keywords:
  - Forecasting
  - Pharmaceutical supply chain
  - Domain knowlege
  - Forecast accuracy
  - Developing countries
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  pdf-engine: pdflatex
  prefer-html: true
  cache: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: International Journal of Production Research
      formatting: preprint
      model: 3p
      cite-style: authoryear
---
